```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# repos="https://utstat.toronto.edu/cran/"
library(tidyverse)
library(stringr)
library(lubridate)
library(maps)
library(dplyr)
```

```{r results='hide', warning=FALSE,}
covid_cases <- read_csv("https://covid.ourworldindata.org/data/owid-covid-data.csv",
                        col_types = cols(
                          .default = col_double(),
                          date = col_date(format =""),
                          location = col_character()
                        ))
covid_cases %>% head()

covid_vaccs <- read_csv("https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv")
covid_vaccs %>% head()

covid_tests <- read_csv("https://covid.ourworldindata.org/data/testing/covid-testing-all-observations.csv")
covid_tests %>% head()

covid_response <- read_csv("https://raw.githubusercontent.com/OxCGRT/covid-policy-tracker/master/data/OxCGRT_latest.csv",
                           col_types = cols(
                             .default = col_character(),
                             Date = col_date(format = "%Y%m%d"))
                           ) %>%
    mutate(
      across(starts_with("Confirmed"), as.integer),
      across(ends_with("Flag"), as.logical), 
      across(ends_with("Index"), as.double), 
      across(ends_with("ForDisplay"), as.double),
      )
covid_response %>% head()

regionclassification <- read_tsv("WorldRegions.csv",skip = 1)
regionclassification %>% head()

area <- read_tsv("AREA.csv", col_names = c("Country", "Area", "ranking", "reigon"),skip = 1)
area %>% head()

population <- read_tsv("POPULATION.csv", col_names = c("Country", "Population", "ranking", "reigon"),skip = 1)
population %>% head()

populationdistribution <- read_tsv("POPULATIONDISTRIBUTION.csv")
populationdistribution %>% head()

urbanization <- read_tsv("URBANIZATION.csv")
urbanization %>% head()
```

 - Tidying Data(idk they all look tidy)

```{r results = 'hide'}
covid_cases
covid_response
covid_tests
covid_vaccs
regionclassification
area
population
populationdistribution
urbanization
```

```{r results = 'hide'}
urbanization_tidy <- urbanization %>%
  select(`Country Name`, `Country Code`, `2020`) %>%
  rename("Urbanization Level" = `2020`)
```






 - Date Selection
 
 So to pick our dates we wanted to see the difference between the peaks of the 2 waves of covid-19. To do this we looked plotted the data set and ploted the amount of new cases each day of 2021
 
```{r}
cases_per_date <- covid_cases %>%
  filter(date > "2021-01-01") %>%
  select(date, new_cases) %>%
  group_by(date)

```
```{r}
cases_per_date %>%
  ggplot(mapping = aes(x = date, y = new_cases)) +
  geom_point() +
  geom_smooth()
```

As we can see there are 2 clear waves of covid where daily cases rise then fall, The 2 happen around mid may and late august.

To find the dates we wanted to use we spit the year into 2 sections before july 1st and after july first as you can see in the next graphs this clearly defines the 2 waves

```{r}
cases_per_date %>%
  filter(date < "2021-07-01") %>%
  ggplot(mapping = aes(x = date, y = new_cases)) +
  geom_point() +
  geom_smooth()
cases_per_date %>%
  filter(date > "2021-07-01") %>%
  ggplot(mapping = aes(x = date, y = new_cases)) +
  geom_point() +
  geom_smooth()
```

Now we filtered for the days with the most new cases in 2021 before july 1st and after july first giving us our dates.

April 28th and august 6th

```{r}
cases_per_date %>%
  filter(date < "2021-07-01") %>%
  arrange(desc(new_cases))
cases_per_date %>%
  filter(date > "2021-07-01") %>%
  arrange(desc(new_cases))
```













 - analysing urbanization
 
Now we wanted to look at the relation urbanization has on covid vacinations, test and cases

```{r}
covid_cases_date1 <- covid_cases %>%
  filter(date == "2021-04-28")
covid_cases_date2 <- covid_cases %>% 
  filter(date == "2021-08-06")
covid_vaccs_date1 <- covid_vaccs %>% 
  filter(date == "2021-04-28")
covid_vaccs_date2 <- covid_vaccs %>%
    filter(date == "2021-08-06")
covid_tests_date1 <- covid_tests %>% 
  filter(Date == "2021-04-28")
covid_tests_date2 <- covid_tests %>%
    filter(Date == "2021-08-06")
```

first we looked at cases. We merged the the urbanization with the covid at. Then we made the boxplot so we could get rid of the outliars from our data
```{r}
urbanization_covid_date1 <- urbanization_tidy %>%
  right_join(covid_cases_date1,by = c("Country Name" = "location"))

urbanization_covid_date2 <- urbanization_tidy %>%
  right_join(covid_cases_date2,by = c("Country Name" = "location"))

urbanization_covid_date2 %>% 
  ggplot(mapping = aes(x = `Urbanization Level`, y = new_cases))+
  ggtitle("2021-08-06")+
  geom_boxplot()

urbanization_covid_date1 %>% 
  ggplot(mapping = aes(x = `Urbanization Level`, y =new_cases))+
  ggtitle("2021-04-08")+
  geom_boxplot()
```

as shown in the box plot there are many outliars in this set. We filtered so the plot chart only shows entries be that have below 1.5 million new cases.
```{r}

urbanization_covid_date1 %>%
  filter(`total_cases` < 5000000) %>%
    ggplot(mapping = aes(x = `Urbanization Level`, y = new_cases))+
    ggtitle("2021-04-08")+
    geom_point()+
  geom_smooth(method= "lm")

urbanization_covid_date2 %>%
  filter(`total_cases` < 5000000) %>%
    ggplot(mapping = aes(x = `Urbanization Level`, y = new_cases))+
    ggtitle("2021-08-06")+
    geom_point() +
  geom_smooth(method = "lm")
```
 
 From this plot we can see there clear correlation between cases and urbanization. The more urban a place is there more cases there are.
 Another thing we see is that on march 28th the correlation between new_cases is a lot stronger than on april 23rd
 
 Next we did the same thing with tests and vaccinations to se the correlation

```{r}
urbanization_vacc_date1 <- urbanization_tidy %>%
  right_join(covid_vaccs_date1,by = c("Country Name" = "location")) %>%
    filter(iso_code %>% str_detect("^[^OWID_]"))


urbanization_vacc_date1 %>% 
  ggplot(mapping = aes(x = `Urbanization Level`, y = daily_vaccinations))+
  ggtitle("2021-04-08")+
  geom_boxplot()

urbanization_vacc_date2 <- urbanization_tidy %>%
  right_join(covid_vaccs_date2,by = c("Country Name" = "location")) %>%
    filter(iso_code %>% str_detect("^[^OWID_]"))

urbanization_vacc_date2 %>% 
  ggplot(mapping = aes(x = `Urbanization Level`, y = daily_vaccinations))+
  ggtitle("2021-08-06")+
  geom_boxplot()
```
For the vaccination se there seemed to be a lot less outliers. we decided to plot the places that had less than 6 hundred thousand daily_vaccinations
``` {r}
urbanization_vacc_date1 %>% 
  filter(daily_vaccinations < 600000) %>%
  ggplot(mapping = aes(x = `Urbanization Level`, y = daily_vaccinations))+
  ggtitle("2021-04-08")+
  geom_point() +
  geom_smooth(method = "lm")



urbanization_vacc_date2 %>% 
  filter(daily_vaccinations < 600000) %>%
  ggplot(mapping = aes(x = `Urbanization Level`, y = daily_vaccinations))+
  ggtitle("2021-08-06")+
  geom_point() +
  geom_smooth(method = "lm")
```
Here we some thing interesting once again in march 8th the relation is a lot stronger than on april 28th. On april 28th the chart seems to be a lot more scattered with more randomness than the first date.


```{r}
urbanization_test_date1 <- urbanization_tidy %>%
  right_join(covid_tests_date1, by = c("Country Code" = "ISO code"))%>%
  arrange(desc(`Daily change in cumulative total`))

urbanization_test_date2 <- urbanization_tidy %>%
  right_join(covid_tests_date2, by = c("Country Code" = "ISO code"))%>%
  arrange(desc(`Daily change in cumulative total`))
urbanization_test_date1 %>%
  ggplot(mapping = aes(x = `Urbanization Level`, y = `Daily change in cumulative total`))+
  geom_boxplot()
urbanization_test_date1 %>%
  filter(`Daily change in cumulative total` < 500000) %>%
  ggplot(mapping = aes(x = `Urbanization Level`, y = `Daily change in cumulative total`))+
  geom_point()
urbanization_test_date2 %>%
    filter(`Daily change in cumulative total` < 500000) %>%
    ggplot(mapping = aes(x = `Urbanization Level`, y = `Daily change in cumulative total`))+
  geom_point()


```
 

 ~~ Urbanization daily vaccinations top 20

Now we decided to look at the places with the most vaccinations on our given date and see how there urbanization varies. 
```{r }
urb_vacc_date1 <- urbanization_vacc_date1 %>%
  filter(is.na(`Country Code`) == FALSE) %>%
  arrange(desc(daily_vaccinations)) %>%
  slice(1:20) %>%
  select("Country Name", "Country Code", "date", "daily_vaccinations", "Urbanization Level")
urb_vacc_date2 <- urbanization_vacc_date2 %>%
  filter(is.na(`Country Code`) == FALSE) %>%
  arrange(desc(daily_vaccinations)) %>%
  slice(1:20) %>%
  select("Country Name", "Country Code", "date", "daily_vaccinations", "Urbanization Level")

urbanization_vaccs_dates <- urb_vacc_date1 %>%
  full_join(urb_vacc_date2) %>%
  mutate(date = as.character(date))

```

```{r}
urbanization_vaccs_dates %>% 
  ggplot()+
  geom_point(mapping = aes(x = daily_vaccinations, y = `Urbanization Level`, colour = date))
  
```
As we can see the urbanization levels vary between the dates. on the march 4th the urbanization level is more top heavy. As for on April 06 the urbanication level is more evenly distributed.

To summarise On march 28th the vaccinations and cases where more linear and top heavy, while on April 6th there was more fluctuation. 

One theory we had of why the data is behaving this way is because More urban places tend to be more populated and richer, making them more in demand for vaccines and more likley to get cases making the data left skewed. 

The reason that the cases and vaccinations are more fluxuant on the second date is because it is later on in the covid cycle and the urban places where have already been vaccinated and are now giving vaccinations to less popluated and less wealthy places.

The reason that the first date is more linear is because these wealther more populated places are still consuming all the vaccinations and are getting more cases because the popuation is higher.

To prove this hypothesis we can look into the population denisity by taking a population and dividing it by area and compare it with both the urbanziation.

```{r}
population
area
pop_density <- population %>%
  left_join(area, by = c("Country")) %>%
  mutate("population_density" = Population/Area)
pop_density
```

```{r}

pop_density %>%
  left_join(urbanization_dates, by = c("Country" = "Country Name")) %>%
  filter(date == "2021-04-28") %>%
  ggplot(mapping = aes(y = population_density, x = `Urbanization Level`), na.rm = TRUE)+
  geom_point()+
  geom_smooth(method = "lm")


pop_density %>%
  left_join(urbanization_dates, by = c("Country" = "Country Name")) %>%
  filter(date == "2021-08-06") %>%
  ggplot(mapping = aes(y = population_density, x = `Urbanization Level`), na.rm = TRUE)+
  geom_point()+
  geom_smooth(method = "lm")

  
```

When looking at the population density verses the urbanization level it disproves our hypothesis as the places with less urbanization tend to have a larger population density. We would be able to look into the hypothesis more if there was a data set on the economy of a country but there is not.

~~ looking at stats by class

One thing we noticed while looking through the vaccine and cases files is that there are there is information given by class. In this part of the report we will be looking at the variance in cases and vaccines in different classes.


```{r}
urbanization_vacc_date1_OWID <- urbanization_tidy %>%
  right_join(covid_vaccs_date1,by = c("Country Name" = "location")) %>%
  filter(`Country Name`== "High income"|`Country Name`== "Upper middle income"|`Country Name`== "Lower middle income"|`Country Name` == "Low income")

urbanization_vacc_date2_OWID <- urbanization_tidy %>%
  right_join(covid_vaccs_date2,by = c("Country Name" = "location")) %>%
  filter(`Country Name`== "High income"|`Country Name`== "Upper middle income"|`Country Name`== "Lower middle income"|`Country Name` == "Low income")
```

To look into the variance between classes we decided to make a count chart showiing the difference in vaccination, cases between the class on each day.

```{r}
urbanization_vacc_date1_OWID %>% 
  full_join(urbanization_vacc_date2_OWID) %>%
    mutate(date = as.character(date)) %>%
  ggplot()+
  geom_count(mapping = aes(x = daily_vaccinations,y = `Country Name`, color = date))


urbanization_cases_date1_OWID <- urbanization_tidy %>%
  right_join(covid_cases_date1,by = c("Country Name" = "location")) %>%
  filter(`Country Name`== "High income"|`Country Name`== "Upper middle income"|`Country Name`== "Lower middle income"|`Country Name` == "Low income")
urbanization_cases_date2_OWID <- urbanization_tidy %>%
  right_join(covid_cases_date2,by = c("Country Name" = "location")) %>%
  filter(`Country Name`== "High income"|`Country Name`== "Upper middle income"|`Country Name`== "Lower middle income"|`Country Name` == "Low income")

urbanization_cases_date1_OWID %>% 
  full_join(urbanization_cases_date2_OWID) %>%
    mutate(date = as.character(date)) %>%
  ggplot()+
  geom_count(mapping = aes(x = new_cases,y = `Country Name`, color = date))

urbanization_cases_date1_OWID %>% 
  full_join(urbanization_cases_date2_OWID) %>%
    mutate(date = as.character(date)) %>%
  ggplot()+
  geom_count(mapping = aes(x = total_cases,y = `Country Name`, color = date))

```

when looking at these graphs the changes are very interesting. on the first daily vaccinations all of the middle class getting vaccinated at a way higher rate then the first date. Also lookin at both dates the low income population is being vaccinated at a way lower rate than everyone else.

Then when looking at the amount of cases on each day respectively. The upper middle class has around the same amount of cases on each day yet they are being vaccinated at a significantly higher rate than the first day.

One thing that may be making the data inconsistent is the poopulation of each class is different. If a certain class has more population they will have more cases and vaccinations

```{r}
urbanization_cases_date1_OWID %>%
  select(-(iso_code)) %>%
  left_join(urbanization_vacc_date1_OWID, be = "date") %>%
  ggplot()+
  geom_point(mapping = aes(x = population, y = daily_vaccinations, color = `Country Name`))

urbanization_cases_date2_OWID %>%
  select(-(iso_code)) %>%
  left_join(urbanization_vacc_date2_OWID, be = "date") %>%
  ggplot()+
  geom_point(mapping = aes(x = population, y = daily_vaccinations, color = `Country Name`))
```
using these graphs we can see in the chart of population vs daily vaccinations that almost stays the same between both dates (of course on date is scaled higher than the other) except that the vaccinations in the higher class sky rockets

```{r}
urbanization_cases_date1_OWID %>%
  ggplot()+
  geom_point(mapping = aes(x = population, y = total_cases, color = `Country Name`))

urbanization_cases_date2_OWID %>%
  ggplot()+
  geom_point(mapping = aes(x = population, y = total_cases, color = `Country Name`))
```

Here we see that the high income is has a very high amount of cases even though they do not make up that much of the population. we decided to calculate the percent of new cases, percent of total cases and percent of population each class takes up
This also justifys some of the other data we saw. for instance the reason that the low income population is geting less vaccinations and less cases and a overall lower rate is because they take up a very little amount of the population compared to the other classes

```{r}

urbanization_cases_date1_OWID %>%
  mutate("pertot" = (total_cases/sum(total_cases)*100), "pernew" = (new_cases/sum(new_cases)*100), "perpop" = (population/sum(population)*100)) %>%
  select(`Country Name`, "pertot", "pernew", "perpop")
urbanization_cases_date2_OWID %>%
  mutate("pertot" = (total_cases/sum(total_cases)*100), "pernew" = (new_cases/sum(new_cases)*100), "perpop" = (population/sum(population)*100)) %>%
  select(`Country Name`, "pertot", "pernew", "perpop")
```

From this calculation we see that the upper class is only  15.7% of the population and takes up over 40 percent of the total cases on both dates. That means that almost half of all covid cases are coming for 15% of the population.
We can see that they are working to fix this problem as the high income population is getting vaccinated at double the rate they are in the first date.

```{r}
urbanization_cases_date1_OWID %>%
  select(-(iso_code)) %>%
  left_join(urbanization_vacc_date1_OWID, be = "date") %>%
  mutate("pertot" = (total_vaccinations/sum(total_vaccinations)*100), "pernew" = (daily_vaccinations/sum(daily_vaccinations)*100), "perpop" = (population/sum(population)*100)) %>%
  select(`Country Name`, "pertot", "pernew", "perpop")
urbanization_cases_date2_OWID %>%
  select(-(iso_code)) %>%
  left_join(urbanization_vacc_date2_OWID, be = "date") %>%
  mutate("pertot" = (total_vaccinations/sum(total_vaccinations)*100), "pernew" = (new_vaccinations/sum(daily_vaccinations)*100), "perpop" = (population/sum(population)*100)) %>%
  select(`Country Name`, "pertot", "pernew", "perpop")
```

The total vaccinations between both dates is also interesting. As we can see high income once again has a very high rate of vaccination and total vaccination considering its low population percentage. In the second date we can see that the percent of total and new vaccinations have gone down in the high income population. This is probably because they got vaccinated early so other classes are catching up.

Another thing that we can see is at how much of a higher rate the upper income classes gets vaccinated than the lower classes. in the first date the higher income groups (high income and upper middle income) make up more then 80 percent of the vaccinated population and more than 80 percent of the people getting vaccinated on the particular day. whilst having the almost the same amount of population percentage

This is concerning as it shows that the lower classes are not getting vaccinated.
